Fractional Types

### 2016-12-18 typespp.tex
did all updates to sec 4

\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\jdg}[3]{{#1} #3}
\newcommand{\evalone}[2]{#1~\triangleright~#2}
\newcommand{\evaloneb}[2]{#1~\triangleleft~#2}
\newcommand{\unitv}{()}
\newcommand{\synch}{\mathsf{{synch}}}
\newcommand{\unitepl}{\mathsf{{unite_+l}}}
\newcommand{\unitipl}{\mathsf{{uniti_+l}}}
\newcommand{\unitepr}{\mathsf{{unite_+r}}}
\newcommand{\unitipr}{\mathsf{{uniti_+r}}}
\newcommand{\swapp}{\mathsf{{swap_+}}}
\newcommand{\assoclp}{\mathsf{{assocl_+}}}
\newcommand{\assocrp}{\mathsf{{assocr_+}}}
\newcommand{\unitetl}{\mathsf{{unite_{\star}l}}}
\newcommand{\unititl}{\mathsf{{uniti_{\star}l}}}
\newcommand{\unitetr}{\mathsf{{unite_{\star}r}}}
\newcommand{\unititr}{\mathsf{{uniti_{\star}r}}}
\newcommand{\swapt}{\mathsf{{swap_{\star}}}}
\newcommand{\assoclt}{\mathsf{{assocl_{\star}}}}
\newcommand{\assocrt}{\mathsf{{assocr_{\star}}}}
\newcommand{\absorbr}{\mathsf{{absorbr}}}
\newcommand{\absorbl}{\mathsf{{absorbl}}}
\newcommand{\factorzr}{\mathsf{{factorzr}}}
\newcommand{\factorzl}{\mathsf{{factorzl}}}
\newcommand{\dist}{\mathsf{{dist}}}
\newcommand{\factor}{\mathsf{{factor}}}
\newcommand{\distl}{\mathsf{{distl}}}
\newcommand{\factorl}{\mathsf{{factorl}}}
\newcommand{\idiso}{\mathsf{{id}}}

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   \forall\exists{\overline{k}}.[\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& \forall\exists{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   \forall\exists{\overline{k}}.[\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& \forall\exists{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  \forall\exists{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& \forall\exists{\overline{k}}.
               [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      \forall\exists{\overline{k}}.
               [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
  & : \epsilon- \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
  & : \epsilon+
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ : \forall\exists{\overline{k}}.[\mathbb{T},v] \leftrightblackspoon
   \forall\exists{\overline{k}}.[\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2] \leftrightblackspoon
  \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
 \forall\exists{\overline{k_3}}.[\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
 \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
 \forall\exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  \forall\exists{\overline{k_1k_3}}.[\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  \forall\exists{\overline{k_2k_4}}.[\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
% \\
% \\
% \Rule{}
% {\jdg{}{}{}}
% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
%   \leftrightblackspoon
%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
% {}
\end{array}\]
\\
Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
\forall\exists{k}.\mathbb{T_2}_\bullet$ uses the $\forall$ in the left-to-right
direction and uses the $\exists$ in the right-to-left direction.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

A type $\mathbb{T}_\bullet$ is generally of the form
$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
lifted to the top leaving a non-dependent type and a value in that type. The
combinators in Fig.~\ref{pifracd:comb} also serve as the operational semantics
for the language as they explain how the value currently in focus changes during
evaluation.

Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
backwards directions where we diambiguate the uses of $\forall$ and $\exists$
for clarity.

\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
%% &\stackrel{\synch_i}{\leftrightblackspoon}&
&k\mapsto i&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\exists&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}, c^i]
\end{array}\]

\[\begin{array}{rcl}
  [\order{c}, c^i]
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^k \rangle,c^k \rangle] \\
&k\mapsto i&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\exists&
  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{c},c^i]
\end{array}\]

Rules for moving quantifiers?

Combinators for packing (creating an existential) and synchronization
(instantiating a universal)??

>>>>

\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\jdg}[3]{{#1} #3}
\newcommand{\evalone}[2]{#1~\triangleright~#2}
\newcommand{\evaloneb}[2]{#1~\triangleleft~#2}
\newcommand{\unitv}{()}
\newcommand{\synchl}{\mathsf{{synchl}}}
\newcommand{\synchr}{\mathsf{{synchr}}}
\newcommand{\packl}{\mathsf{{packl}}}
\newcommand{\packr}{\mathsf{{packr}}}
\newcommand{\unitepl}{\mathsf{{unite_+l}}}
\newcommand{\unitipl}{\mathsf{{uniti_+l}}}
\newcommand{\unitepr}{\mathsf{{unite_+r}}}
\newcommand{\unitipr}{\mathsf{{uniti_+r}}}
\newcommand{\swapp}{\mathsf{{swap_+}}}
\newcommand{\assoclp}{\mathsf{{assocl_+}}}
\newcommand{\assocrp}{\mathsf{{assocr_+}}}
\newcommand{\unitetl}{\mathsf{{unite_{\star}l}}}
\newcommand{\unititl}{\mathsf{{uniti_{\star}l}}}
\newcommand{\unitetr}{\mathsf{{unite_{\star}r}}}
\newcommand{\unititr}{\mathsf{{uniti_{\star}r}}}
\newcommand{\swapt}{\mathsf{{swap_{\star}}}}
\newcommand{\assoclt}{\mathsf{{assocl_{\star}}}}
\newcommand{\assocrt}{\mathsf{{assocr_{\star}}}}
\newcommand{\absorbr}{\mathsf{{absorbr}}}
\newcommand{\absorbl}{\mathsf{{absorbl}}}
\newcommand{\factorzr}{\mathsf{{factorzr}}}
\newcommand{\factorzl}{\mathsf{{factorzl}}}
\newcommand{\dist}{\mathsf{{dist}}}
\newcommand{\factor}{\mathsf{{factor}}}
\newcommand{\distl}{\mathsf{{distl}}}
\newcommand{\factorl}{\mathsf{{factorl}}}
\newcommand{\idiso}{\mathsf{{id}}}

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   \forall\exists{\overline{k}}.[\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& \forall\exists{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   \forall\exists{\overline{k}}.[\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& \forall\exists{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  \forall\exists{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& \forall\exists{\overline{k}}.
               [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      \forall\exists{\overline{k}}.
               [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
  & : \epsilon- \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
  & : \epsilon+ \\
\synchl :& \forall\exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle, c^i]
  & \leftrightblackspoon &
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
    \langle \langle c^i,1/c^i \rangle, c^i]
  & : \packl \\
\synchr :& \forall\exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k, c^k \rangle \rangle]
  & \leftrightblackspoon &
   [\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^i, \langle 1/c^i, c^i \rangle \rangle]
  & : \packr
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ : \forall\exists{\overline{k}}.[\mathbb{T},v] \leftrightblackspoon
   \forall\exists{\overline{k}}.[\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2] \leftrightblackspoon
  \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
 \forall\exists{\overline{k_3}}.[\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
 \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
 \forall\exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  \forall\exists{\overline{k_1k_3}}.[\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  \forall\exists{\overline{k_2k_4}}.[\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
\end{array}\]
\\
Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
\forall\exists{k}.\mathbb{T_2}_\bullet$ introduces a $\forall$ in the left-to-right
direction and eliminates an $\exists$ in the right-to-left direction. A rule
$\forall\exists{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
\mathbb{T_2}_\bullet$ eliminates a $\forall$ in the left-to-right direction and
introduces an $\exists$ in the right-to-left direction.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

A type $\mathbb{T}_\bullet$ is generally of the form
$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
lifted to the top leaving a non-dependent type and a value in that type. The
combinators in Fig.~\ref{pifracd:comb} also serve as the operational semantics
for the language as they explain how the value currently in focus changes during
evaluation. Given the new combinators, we can modify the the circuit in
Fig.~\ref{fig:zigzag} by adding explicit synchronization primitives. The
evaluation proceeds as follows where we diambiguate the uses of $\forall$ and
$\exists$ for clarity.

Forward execution:
\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\synchl}{\leftrightblackspoon}&
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\stackrel{\packr}{\leftrightblackspoon}&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}, c^i]
\end{array}\]

Reverse execution:
\[\begin{array}{rcl}
  [\order{c}, c^i]
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\synchr}{\leftrightblackspoon}&
 [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\stackrel{\packl}{\leftrightblackspoon}&
  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{c},c^i]
\end{array}\]
